---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tlda

<!-- badges: start -->
<!-- badges: end -->

This package includes a number of utility functions and resources for language data analysis, with a focus on corpus-linguistic research tasks.


## Installation

You can install the development version of tlda from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("lsoenning/tlda")
```

## Example: Dispersion measures

The following example demonstrates the use of the function `disp()` for calculating various dispersion measures based on two vectors:

- `subfreq` a set of subfrequencies, i.e. the number of occurrences of the item in each corpus part
- `partsize` a vector with the size of the corpus parts

The argument `directionality` controls the scaling of the scores:

- `conventional`: higher values reflect a **more even** distribution
- `gries`: higher values reflect a **less even** distribution

The function prints information about the directionality of scaling and details about the formula used.

As an example, we will use data from Lyne's (1985) classic study and consider the distribution of the French lemma ALLEMAND across the ten (nearly) equal-sized parts ('Tenths') of his corpus of French business correspondence. The part sizes are taken from Figure 1 (p. 85) and the subfrequencies from Appendix I (p. 299). The following dispersion measures are calculated (returned in chronological order):

- $R_{rel}$ relative range (Keniston 1920)
- $D$ (Juilland & Chang-Rodriguez 1964)
- $D_2$ (Carroll 1970)
- $S$ (Rosengren 1971)
- $D_P$ (Gries 2008; modification: Egbert et al. 2020)
- $D_A$ (Burch et al. 2017)
- $D_{KL}$ (Gries 2020, 2021)

```{r}
library(tlda)

x <- c(2, 0, 1, 1, 3, 0, 3, 0, 0, 0)
y <- c(8143, 8058, 8271, 8125, 7959, 7941, 8146, 8001, 8003, 7930)

disp(
  subfreq = x,
  partsize = y,
  directionality = "conventional"
)
```

If we prefer the reversed scaling used by Gries (2008), we can change the value of the argument `directionality`, like so:

```{r}
disp(
  subfreq = x,
  partsize = y,
  directionality = "gries"
)
```

Critically, the functions offered by the current package for calculating dispersion **keep apart the formula and the directionality of scaling**. This means that the `directionality` argument always overrides the scaling implemented in the original formula; the default is `conventional`.

For three dispersion measures, separate functions offer more options:

- `disp_R()` for $Range$
- `disp_DP()` for Gries's *deviation of proportions*
- `disp_DA()` for $D_A$

For $Range$, three choice are available: (i) relative range (`relative`), i.e. the *proportion* of corpus parts containing at least one occurrence of the item (default); (ii) absolute range (`absolute`), i.e. the *number* of corpus parts containing at least one occurrence of the item; and (iii) relative range with size (`relative_withsize`), the proportional version that takes into account the size of the corpus parts (see Gries 2022: 179-180; Gries 2024: 27-28). The following code returns absolute range:

```{r}
disp_R(
  subfreq = x,
  partsize = y,
  type = "absolute"
)
```

For Gries's *deviation of proportions*, the function `disp_DP()` allows the user to choose from three different formulas that are found in the literature. The following code uses the original formula in Gries (2008), but with `conventional` scaling:

```{r}
disp_DP(
  subfreq = x,
  partsize = y,
  formula = "gries_2008",
  directionality = "conventional"
)
```

For $D_A$, three formulas are available, The `basic` version is direct implementation of the actual formula for this measure, which is computationally expensive if the number of corpus parts is large. Wilcox (1973: 343) gives a `shortcut` version, which is much quicker (see this [blog post](https://lsoenning.github.io/posts/2023-12-11_computation_DA/)). Finally, `shortcut_mod` is a slightly adapted form of the shortcut (experimental), which ensures that scores do not exceed 1 (conventional scaling). The following code uses Wilcox's (1973) shortcut version:

```{r}
disp_DA(
  subfreq = x,
  partsize = y,
  formula = "shortcut",
  directionality = "conventional"
)
```

## References

Burch, Brent, Jesse Egbert & Douglas Biber. 2017. Measuring and interpreting lexical dispersion in corpus linguistics. \emph{Journal of Research Design and Statistics in Linguistics and Communication Science} 3(2). 189–216. \url{https://doi.org/10.1558/jrds.33066}

Carroll, John B. 1970. An alternative to Juilland’s usage coefficient for lexical frequencies and a proposal for a standard frequency index. \emph{Computer Studies in the Humanities and Verbal Behaviour} 3(2). 61–65. \url{https://doi.org/10.1002/j.2333-8504.1970.tb00778.x}

Egbert, Jesse, Brent Burch & Douglas Biber. 2020. Lexical dispersion and corpus design. \emph{International Journal of Corpus Linguistics} 25(1). 89–115. \url{https://doi.org/10.1075/ijcl.18010.egb}

Gries, Stefan Th. 2008. Dispersions and adjusted frequencies in corpora. \emph{International Journal of Corpus Linguistics} 13(4). 403–437. \url{https://doi.org/10.1075/ijcl.13.4.02gri}

Gries, Stefan Th. 2020. Analyzing dispersion. In Magali Paquot & Stefan Th. Gries (eds.), \emph{A practical handbook of corpus linguistics}, 99–118. New York: Springer. \url{https://doi.org/10.1007/978-3-030-46216-1_5}

Gries, Stefan Th. 2021. A new approach to (key) keywords analysis: Using frequency, and now also dispersion. \emph{Research in Corpus Linguistics} 9(2). 1−33. \url{https://doi.org/10.32714/ricl.09.02.02}

Juilland, Alphonse G. & Eugenio Chang-Rodríguez. 1964. \emph{Frequency dictionary of Spanish words.} The Hague: Mouton de Gruyter. \url{https://doi.org/10.1515/9783112415467}

Keniston, Hayward. 1920. Common words in Spanish. \emph{Hispania} 3(2). 85–96. \url{https://doi.org/10.2307/331305}

Lijffijt, Jefrey & Stefan Th. Gries. 2012. Correction to Stefan Th. Gries’ ‘Dispersions and adjusted frequencies in corpora’. \emph{International Journal of Corpus Linguistics} 17(1). 147–149. \url{https://doi.org/10.1075/ijcl.17.1.08lij}

Lyne, Anthony A. 1985. \emp{The vocabulary of French business correspondence}. Paris: Slatkine-Champion.

Rosengren, Inger. 1971. The quantitative concept of language and its relation to the structure of frequency dictionaries. \emph{Études de linguistique appliquée (Nouvelle Série)} 1. 103–127.